---
title: "SVA Simulation"
author: "Christopher Lo"
date: "9/6/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo=F, message=F, warning=F}
library(sva)
library(tidyverse)
library(invgamma)
```

# Simulation Set-Up

Simulation studies recreated to my best abilities from 
"A general framework for multiple testing dependence" (Leek et al. 2008)

We generate $X$ from the following model: $X = BS + \Gamma G + U$. 

We have 1000 genes, 20 samples, and 2 latent variables. The design matrix $S$ is a case control set-up of 10 cases and 10 controls. 

$$\sigma_i^2 \sim InvGamma(10, 9)$$

$$U_{ij} \sim N(0, \sigma_i^2)$$
$$b_{m,1} \sim N(0, 1), m=1,...,1000$$
$$b_{m, 2} \sim N(0, 3.1), m=1,...,300$$
$$b_{m, 2} \sim N(2.5, 1), m=301,...,1000$$

$$\Gamma_{m, 1} \sim N(0, 2.5), m=301,...,700$$
$$\Gamma_{m, 2} \sim N(0, 2.5), m=501,...,900$$

$$G_{1:10} \sim Bernoulli(.7)$$
$$G_{11:20} \sim Bernoulli(.2)$$
```{r echo=F}
m = 1000
n = 20
r = 2

sigma_i = rinvgamma(m, 10, 9)
U = t(sapply(sigma_i, function(x) rnorm(20, mean = 0, sd = x)))

b1 = rnorm(m, mean = 0, sd = 1)
b2 = mapply(mu = c(rep(0, 300), rep(3.1, m - 300)),
             sigma = c(rep(2.5, 300), rep(1, m - 300)),
             function(mu, sigma) rnorm(1, mean = mu, sd = sigma))
B = cbind(b1, b2)
S = matrix(c(rep(1, n), rep(0, 10), rep(1, 10)), byrow = T, ncol = n)

Gamma = t(mapply(mu = c(rep(0, m)),
                 sigma = c(rep(2.5, m)),
                 function(mu, sigma) rnorm(r, mean = mu, sd = sigma)))
Gamma[c(1:200, 701:m), 1] = 0 #only signal from genes 300-700 in SV1
Gamma[c(1:400, 901:m), 2] = 0 #only signal from genes 500-900 in SV2
G = mapply(p = c(rep(.7, 10), rep(.2, 10)), #no need for intercept for kernel G. 
           function(p) rbernoulli(2, p))


X = B %*% S + Gamma %*% G + U
```

# Run SVA and regression to estimate parameters and SVs

We look at the number of SVs estimated, whether the latent variables are spanned by the estimated SVs, the recovered regression coefficients, the null p-value distribution, and the ranks of top genes. 

Todo: ranks of top genes code unsure right now. 


```{r echo=T, warning=F, message=F}
nullMod = t(S)[, 1]
n.sv = num.sv(X, t(S), method = "be")
cat("Method be: Number of SVs: ", n.sv, "\n")

leek.n.sv = num.sv(X, t(S), method = "leek")
cat("Method leek: Number of SVs: ", leek.n.sv, "\n")

svobj = sva(X, t(S), nullMod, n.sv = n.sv)

#visually look at predicted SVs. 
qplot(as.numeric(G[1 ,]), svobj$sv[, 1], xlab = "True latent variable 1", ylab = "Est. SV1")
qplot(as.numeric(G[2 ,]), svobj$sv[, 2], xlab = "True latent variable 2", ylab = "Est. SV2")

nullmodsv = cbind(nullMod, svobj$sv)
modsv = cbind(t(S), svobj$sv)
#run full regression.
fitsv = lm.fit(modsv, t(X))

#visually look at predicted coefficients
effect = c(rep("null", 300), rep("alt", m - 300))
qplot(B[, 1], fitsv$coefficients[1 ,], color = effect, main = round(cor(fitsv$coefficients[1 ,], B[, 1]), 2), alpha = .4, xlab = "True b_i1", ylab = "Est. b_i1")
qplot(B[, 2], fitsv$coefficients[2 ,], color = effect, main = round(cor(B[, 2], fitsv$coefficients[2 ,]), 2), alpha = .25, xlab = "True b_i2", ylab = "Est. b_i2")

effect = c(rep("null", 200), rep("alt", 500), rep("null", 300))
qplot(Gamma[, 1], fitsv$coefficients[3 ,], color = effect, main = round(cor(Gamma[, 1], fitsv$coefficients[3 ,]), 2), alpha = .4, xlab = "True gamma_i1", ylab = "Est. gamma_i1")

effect = c(rep("null", 400), rep("alt", 500), rep("null", 100))
qplot(Gamma[, 2], fitsv$coefficients[4 ,], color = effect, main = round(cor(Gamma[, 2], fitsv$coefficients[4 ,]), 2), alpha = .4, xlab = "True gamma_i2", ylab = "Est. gamma_i2")

#just compute p-value of b_12 = 0 using F statistics. 
pValuesSv = f.pvalue(X, modsv, nullmodsv)
#double check with existing function:
pValuesSV2 = rep(NA, 1000)
fstat = rep(NA, 1000)
for(i in 1:1000) { 
  dat = data.frame(x = X[i ,], pv = S[2 ,], sv1 = svobj$sv[, 1], sv2 = svobj$sv[, 2])
  dat_nullmod = lm(x ~ pv, dat)
  dat_mod = lm(x ~ pv + sv1 + sv2, dat)
  an = anova(dat_nullmod, dat_mod)
  pValuesSV2[i] = an$`Pr(>F)`[2]
  fstat[i] = an$F[2]
}


qValuesSv = p.adjust(pValuesSv, method = "BH")
qValuesSV2 = p.adjust(pValuesSV2, method = "BH")


plot(1:length(qValuesSV2[1:300])/(length(qValuesSV2[1:300])+1),sort(qValuesSV2[1:300]))
abline(a = 0, b = 1)
ks.test(qValuesSV2[1:300],"punif",0,1)

table(qValuesSV2[1:300] < .05)
table(qValuesSV2[301:1000] < .05)

fstat_rank = rank(-fstat)
true_rank = rank(-abs(b2))

qplot(true_rank, fstat_rank)
```

