---
title: "SVA Recount"
author: "Christopher Lo"
date: "9/19/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo=F, message=F, warning=F}
library(tidyverse)
library(recount3)
library(sva)
library(DESeq2)
library(ggfortify)
theme_set(theme_bw())
```


## Introduction

*What happens to the number of Surrogate Variables as we increase the number of studies in recount3?*

To illustrate this, we examined 4 studies that involved case/control prostate tumor and normal samples,
with a minimal of 25 samples per study. For each study, we examined the number of SVs (using method = "be"), 
and we also aggregated them one by one and examined the number of SVs. 

1.    SRP118614: "Overall design: Matched high-grade (GS=7(4+3)) prostate tumor and adjacent
normal specimens from 16 patients (8 AAM and 8 EAM) were subjected to two 
replicate runs of RNA-sequencing."

2.    SRP002628: "Overall design: We sequenced the transcriptome (polyA+) of 20 prostate cancer 
tumors and 10 matched normal tissues using Illumina GAII platform. 
Then we used bioinformatic approaches to identify prostate cancer specific 
aberrations which include gene fusion, alternative splicing, somatic mutation."

3.    SRP212704: "Overall design: Strand specific total RNA seq was performed using frozen 
patient matched prostate cancer tissue in biological duplicates.
Purpose: The goal of present study is to compare transcript level changes
between normal and tumor of same individuals"  

4.    SRP027258: "We utilized RNA sequencing to test the hypothesis that SFN modifies the 
expression of genes that are critical in prostate cancer progression. 
Normal prostate epithelial cells, and androgen-dependent and androgen-independent 
prostate cancer cells were treated with 15 µM SFN and the transcriptome was 
determined at 6 and 24 hour time points."


```{r echo=F, message=F}
#Setup recount3
human_projects <- available_projects()
```

```{r echo=F, message=F}
#Studies we want to look at, and what we should grep for to find the case status
#under "sra.sample_attributes"
study_info = data.frame(studyName = c("SRP118614", "SRP002628", "SRP212704", "SRP027258"),
                        caseStatus_grep = c("prostate tumor", "Prostate cancer tissue", "Tumor", "prostate cancer"))
```


```{r echo=F, message=F}
#Define analysis functions.

analyzeSingleStudy = function(studyName, caseStatus_grep) {
  #Load in phenotype metadata
  proj_info <- subset(human_projects, project == studyName)
  rse_gene <- create_rse(proj_info)
  phenotype = data.frame(colData(rse_gene))
  phenotype$phenotype_tumor = grepl(caseStatus_grep, phenotype$sra.sample_attributes)
  mod = model.matrix(~phenotype_tumor, data = phenotype)
  #Load in raw counts, and nomralize it via sizeFactors from DESeq. 
  geneCounts = assays(rse_gene)$raw_counts
  dds <- DESeqDataSetFromMatrix(countData = geneCounts, colData = phenotype, design = ~ phenotype_tumor)
  dds <- estimateSizeFactors(dds)
  geneCountsNormalized <- counts(dds, normalized = TRUE)
  #Summary statistics of sizeFactors. 
  print("SizeFactor distribution:")
  print(summary(sizeFactors(dds)))
  
  #SVA analysis
  n.sv = num.sv(geneCountsNormalized, mod, method = "be", vfilter = 10000)
  #PCA analysis
  pca = prcomp(t(geneCountsNormalized))
  variance = pca$sdev^2 / sum(pca$sdev^2)
  variance = variance[1:10]
  #PCA plots
  g = qplot(c(1:length(variance)), variance) + geom_line() + geom_point() +
        geom_hline(yintercept = 1/ncol(geneCounts), linetype = "dashed") +
        xlab("Principal Component") + ylab("Variance Explained") + 
        ggtitle(paste0(studyName, "\nNumber of SVs: ", n.sv))  + 
        ylim(0, 1)
  print(g)
  print(autoplot(pca, data = phenotype, colour = 'phenotype_tumor'))

  return(list(n.sv = n.sv,
              geneCounts = geneCounts, #return raw counts. 
              mod = mod))
}
```


```{r echo=F, message=F}
#Define analysis functions.

analyzeMultipleStudies = function(mods, geneCounts, studyNames) {
  stopifnot(class(mods) == "list", class(geneCounts) == "list", class(studyNames) == "character")
  stopifnot(length(mods) > 1, length(geneCounts) > 1, length(studyNames) > 1)
  
  #Get study variable
  studyNamesLong = character()
  for(j in 1:length(studyNames)) {
    studyNamesLong = c(studyNamesLong,
                       rep(studyNames[j], ncol(geneCounts[[j]])))
  }
  
  #Collapse list to dataframe
  geneCounts = do.call(cbind, geneCounts)
  mods = do.call(rbind, mods)
  
  #normalize across studies
  dds <- DESeqDataSetFromMatrix(countData = geneCounts, colData = mods, design = ~ phenotype_tumorTRUE)
  dds <- estimateSizeFactors(dds)
  geneCountsNormalized <- counts(dds, normalized=TRUE)
  
  #SVA analysis
  n.sv = num.sv(geneCountsNormalized, as.matrix(mods), method = "be", vfilter = 10000)
  #PCA analysis
  pca = prcomp(t(geneCountsNormalized))
  variance = pca$sdev^2 / sum(pca$sdev^2)
  variance = variance[1:10]
  #PCA plots
  g = qplot(c(1:length(variance)), variance) + geom_line() + geom_point() + 
        geom_hline(yintercept = 1/ncol(geneCounts), linetype = "dashed") +
        xlab("Principal Component") + ylab("Variance Explained") + 
        ggtitle(paste0(studyNames, collapse = " + "),
                subtitle = paste0("Number of SVs: ", n.sv)) + 
        ylim(0, 1)
  print(g)

  mods = as.data.frame(mods)
  mods$studyName = studyNamesLong
  mods$phenotype_tumorTRUE = as.factor(mods$phenotype_tumorTRUE)
  print(autoplot(pca, data = mods, colour = 'studyName', shape = 'phenotype_tumorTRUE'))
}
```

## Analysis

```{r echo=F, message=F}
#Run analysis here.

agg_mods = list()
agg_geneCounts = list()


for(i in 1:nrow(study_info)) {
  study_i = analyzeSingleStudy(study_info$studyName[i], 
                               study_info$caseStatus_grep[i])
  
  agg_mods[[i]] = study_i$mod
  agg_geneCounts[[i]] = study_i$geneCounts
  
  if(i > 1)
    analyzeMultipleStudies(agg_mods, agg_geneCounts, study_info$studyName[1:i])
}
```

